---
title: "Retirement projections"
author: "Gurbinder Singh"
date: "`r Sys.Date()`"
output: html_document
---

This is an example of a retirement calculator:
https://www.nerdwallet.com/calculator/retirement-calculator#results

```{r}
library(tidyverse)
```

# Setup initial data

## Inflation data

Source: https://www.infina.at/trends/statistik-inflation/.

```{r}
inflation.data <- tibble(
  year = c( 2000, 2001, 2002, 2003, 2004, 2005, 2006, 2007, 2008, 2009, 2010, 
            2011, 2012, 2013, 2014, 2015, 2016, 2017, 2018, 2019, 2020, 2021, 
            2022, 2023),
  inflation = c(2.3, 2.7, 1.8, 1.3, 2.1, 2.3, 1.5, 2.2, 3.2, 0.5, 1.9, 3.3, 
                  2.4, 2.0, 1.7, 0.9, 0.9, 2.1, 2.0, 1.5, 1.4, 2.8, 8.6, 7.8)
  )

average.inflation <- 1 + mean(inflation.data$inflation) / 100
```


## Setup

```{r}
# TODO: add variable and fix cost properties
current.net.salary <- 1500
net.annual.salary <- current.net.salary * 14
savings.percentage <- 0.15
pension.estimate <- 0.72
investment.return.pa <- 1.075
capital.gains.tax <- 0.275
current.savings <- 2000

# Increase ratio relative to inflation.
increase.ratio <- 0.75
average.salary.increase <- 1 + (average.inflation - 1) * increase.ratio

start.year <- 30 
end.year <- 80
pension.start.year <- 65
working.years <- pension.start.year - start.year
total.years <- end.year - start.year

projection <- tibble(
  year = start.year:end.year,
  inflation = rep(average.inflation, total.years + 1) ^ (0:total.years),
  salary = c(rep(net.annual.salary, working.years), 
             rep(0, total.years - working.years + 1))
)

write_csv2(projection, "projections-1.csv")
```


# Projected income

## Salary

Calculate the projected monthly salary assuming we work until the age of 65 and the salary increases at a certain praction of the inflation.

```{r}
projection <-
  projection |>
  mutate(salary = salary * (average.salary.increase ^ (0:total.years)))
```

## Pension

The pension is roughly 80% of the average salary of the 30 highest paid years. Then we calculate the monthly pension up to the age of 90. Again we assume an annual increase of 2/3 of the inflation.

```{r}
expected.starting.pension <-
  (projection |>
     filter(salary > 0) |>
     tail(30))$salary |>
  mean() * pension.estimate

pension.years <- total.years - working.years

projection$pension <- 
  c(rep(0, working.years), 
    expected.starting.pension * (average.salary.increase ^ (0:pension.years)))

write_csv2(projection, "projections-2.csv")
```


# Living costs

For the living costs (based on life style) we assume to be the salary minus the percentage we save.

```{r}
projection$living.costs <- 
  net.annual.salary * (1 - savings.percentage) * projection$inflation
```


# Savings and investments

Calculate the cummulative savings per year as well as if the same amount were invested.

## Savings

```{r}
projection <-
  projection |>
  # mutate(deficit = pmin(0, available.money)) |>
  # mutate(savings.var = pmax(0, available.money), 
  #        savings.fix = salary * savings.percentage)
  mutate(savings.var = salary + pension - living.costs) |>
  mutate(savings.fix = salary * savings.percentage + pmin(0, savings.var))
         

projection$savings.var[1] <- projection$savings.var[1] + current.savings
projection$savings.fix[1] <- projection$savings.fix[1] + current.savings

projection <- 
  projection |>
  mutate(savings.var.cmlt = cumsum(savings.var),
         savings.fix.cmlt = cumsum(savings.fix))

write_csv2(projection, "projections-3.csv")
```

## Investements

### Formula

The equation for calculating the net value of a share is

$$
n = g - (g - o) * t
$$

where `n` is the net value, `g` the gross value, `o` the original value and `t` the tax.

We can use this to derive the inverse:

$$
n = g - \frac{g}{g} * (g-o)*t \\
n = g * (1 - \frac{g-o}{g} * t)
$$

since 

$$
growth = \frac{g}{o} \\
\Rightarrow \  
g = growth * o
$$

we can also write

$$
n = g * (1 - \frac{o*growth - o}{o*growth} * t) \\
n = g * (1 - \frac{growth - 1}{growth} * t) \\
g = \frac{n}{(1 - \frac{growth - 1}{growth} * t)}
$$

### Code

```{r}
gross.value <- function(net.value, growth.factor, tax) {
  profit.margin <- (growth.factor - 1) / growth.factor
  # cat(net.value, profit.margin, tax, "\n")
  net.value / (1 - profit.margin * tax)
}

process.investment <- function(annual.investments) {
  investments.value <- annual.investments
  invested.amount <- annual.investments
  
  # At the start the vector market.value
  for (i in 2:length(annual.investments)) {
    current <- investments.value[i - 1] * investment.return.pa
    base <- invested.amount[i - 1]
    new <- annual.investments[i]
    growth.factor <- current / base
    # cat("invested = ", base,
    #     ", investments =", current,
    #     ", growth =", growth.factor,
    #     ", new investments =", new,
    #     "\n")
    
    if (new >= 0 || current <= 0) {
      investments.value[i] <- pmax(0, current + new)
      invested.amount[i] <- pmax(0, base + new)
      
    } else {
      new <- abs(new)
      to.sell <- gross.value(new, growth.factor, capital.gains.tax)
      # cat("need =", new, ", sell =", to.sell, "\n")
      
      investments.value[i] <- pmax(0, current - to.sell)
      invested.amount[i] <- pmax(0, base - to.sell / growth.factor)
    }
  }
  investments.value
}

projection$invested.var <- process.investment(projection$savings.var)
projection$invested.fix <- process.investment(projection$savings.fix)

write_csv2(projection, "projections-4.csv")
```

# Plot

```{r}
projection |>
  select(year,
         savings.var.cmlt,
         savings.fix.cmlt,
         invested.var,
         invested.fix) |>
  pivot_longer(cols = !c("year"),
               names_to = "type",
               values_to = "amount") |>
  filter(amount > 0) |>
  ggplot(aes(x = year, y = amount, color = type)) +
  geom_line() +
  labs(
    title = "Available capital per year",
    subtitle = paste(
      "Assuming a capital growth of",
      (investment.return.pa * 100 - 100),
      "percent per year and a salary increase of",
      (increase.ratio * 100),
      "percent of the inflation. \nAnd a fixed savings rate of",
      (savings.percentage * 100),
      "percent of the salary (", current.net.salary, 
      "Euro in", start.year, ")"
    ),
    x = "Years",
    y = "Amount"
  ) +
  scale_x_continuous(breaks = seq(start.year, end.year, 10)) +
  scale_y_continuous(n.breaks = 10)

ggsave("available-capital.2.png", scale = 2)
```
